Modelling circumstellar discs with 3D radiation hydrodynamics

### 2009-10-12 SPH-Torus-paper1/sphtorus_benchmark.tex
Minor updates
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2051 25464f21-6099-402c-b30a-640f1d40d38b

The solution is to include radiation-transfer self-consistently  and perform a radiation-hydrodynamics (RHD) calculation. The simplest way to do this is to assume that the energy is transported by radiative diffusion of photons. The diffusion approximation is a good one at high optical depth, but breaks down in optically thin regions since the diffusion speed may exceed the speed of light as the mean-free-path tends to infinity; an essentially arbitrary flux limiter is adopted in this regime. Furthermore the approximation is often implemented as a grey method, typically using Rosseland mean  opacities tabulated as a function of temperature. Finally the diffusion approximation cannot treat the effects of shadowing, as radiation field may diffuse around optically thick obstacles. The crucial advantage of the diffusion approximation though is of course speed: the radiation transport step is straightforward to implement and is not hugely dominant in the RHD calculation. The first cluster collapse and disc fragmentation calculations using flux-limited diffusion have just started appearing in the literature \citep{boss_2008, bate_2009b, offner_2009, stamatellos_2009}. The fundamental result of including a more complete thermal treatment is broadly in line with prior expectations i.e.  the heating of the gas inhibits the fragmentation process.

@@

To illustrate the effects of the different splitting methods two
example grids are shown in Fig.~\ref{fig:example_meshes}. A grid
constructed from $10^7$ particles, using a mass per cell limit of
$5\times10^{26}\,\rm{g}$, is plotted in
Fig.~\ref{fig:mass_condition_mesh}. This plot is a slice perpendicular
to the disc mid-plane and shows increased resolution towards the disc
centre and mid-plane where the density is higher.  A grid with
$f_{\rm{split}}=0.1$, also using $10^7$ particles, is plotted in
Fig.~\ref{fig:density_condition_mesh}. The resolution increases
towards the edge of the disc, where there are large density gradients,
and also increases towards the centre of the disc. 
\begin{figure*}
  \subfigure[AMR grid with a mass per cell condition of
    $5\times10^{26}\,\rm{g}$.]
    {\includegraphics[scale=0.22]{figures/set1_mesh}
  \label{fig:mass_condition_mesh}}
  \subfigure[Grid with a density condition of
    $f_{\rm{split}}=0.1$.]
    {\includegraphics[scale=0.22]{figures/set2_mesh}
      \label{fig:density_condition_mesh}}
  \caption{Example AMR grids for a mass per cell splitting condition
    of  $5\times10^{26}\,\rm{g}$ (left)
    and a density contrast splitting condition of $f_{\rm{split}}=0.1$
    (right) constructed
    from $10^7$ SPH particles. The grids are
    slices perpendicular to the disc mid-plane and the axis units are $10^{16}$\,cm.}
  \label{fig:example_meshes}
\end{figure*}

@@

SEDs were calculated using the {\sc torus} code for viewing
angles of 12.5 and 77.5 degrees (where an inclination angle of zero
degrees indicates that the disc is viewed face-on) and the results are
plotted in Fig.~\ref{fig:seds} (dashed line). Also plotted in
Fig.~\ref{fig:seds} are the benchmark results of \cite{pascucci_2004}
(solid line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/sed_013}
  \includegraphics[scale=0.3]{figures/sed_077}
  \caption{SEDs from the benchmark disc represented by $10^5 $ particles and a
    Cartesian radiative transfer grid generated using a mass per cell
    limit of $5\times10^{26}$\,g. SEDs are shown for inclination
    angles of 12.5 degrees (left) and 77.5 degrees (right). The
    benchmark SED is plotted as a solid line and the {\sc torus} SED
    is plotted as a dashed line. The corresponding fractional
    temperature errors are plotted in Fig.\ref{fig:nomod_tem_diff}.}
  \label{fig:seds}
\end{figure*}
Both SEDs show significant departures from the benchmark results;
there is excess emission between 10-100 $\mu\rm{m}$ and at a viewing
angle of 77.5 degrees there is also a significant reduction in flux
below 1\,$\mu\rm{m}$. Both these discrepancies are attributable to a lack of resolution within a few au of the central object. The density distribution in the central region of
this disc is plotted in Fig.~\ref{fig:basic_dens} which shows that the
central 1\,au gap is not represented. There are two effects in
operation; firstly the grid resolution in this region is larger than
1\,au  and secondly the smoothing lengths of the particles are larger
than 1\,au. In order to correctly represent the central gap there needs
to be higher grid resolution around the central source and more SPH
particles are required so that the smoothing lengths are smaller.
\begin{figure}
\includegraphics[scale=0.2]{figures/basic_dens}
\caption{Density in the central region of the benchmark disc
  represented by $10^5$ particles and a Cartesian radiative transfer
  grid generated using a mass per cell limit of
  $5\times10^{26}$\,g. Axis units are $10^{13}$cm. The central gap of
  1\,au ($1.5\time10^{13}$\,cm) is not represented and the effects on
  the temperature distribution and emergent SEDs are shown in
  Fig.~\ref{fig:nomod_tem_diff} and Fig.~\ref{fig:seds}
  respectively.}
\label{fig:basic_dens}
\end{figure}

>>>>

The solution is to include radiation-transfer self-consistently  and
perform a radiation-hydrodynamics (RHD) calculation. The simplest way
to do this is to assume that the energy is transported by radiative
diffusion of photons. The diffusion approximation is a good one at
high optical depth, but breaks down in optically thin regions since
the diffusion speed may exceed the speed of light as the
mean-free-path tends to infinity; an essentially arbitrary flux
limiter is adopted in this regime. Furthermore the approximation is
often implemented as a grey method, typically using Rosseland mean
opacities tabulated as a function of temperature. Finally the
diffusion approximation cannot treat the effects of shadowing, as the radiation field may diffuse around optically thick obstacles. The crucial advantage of the diffusion approximation though is of course speed: the radiation transport step is straightforward to implement and is not hugely dominant in the RHD calculation. The first cluster collapse and disc fragmentation calculations using flux-limited diffusion have just started appearing in the literature \citep{boss_2008, bate_2009b, offner_2009, stamatellos_2009}. The fundamental result of including a more complete thermal treatment is broadly in line with prior expectations i.e.  the heating of the gas inhibits the fragmentation process.

@@

To illustrate the effects of the different splitting methods two
example grids are shown in Fig.~\ref{fig:example_meshes}. A grid
constructed from $10^7$ particles, using a mass per cell limit of
$5\times10^{26}\,\rm{g}$, is plotted in
Fig.~\ref{fig:mass_condition_mesh}. This plot is a slice perpendicular
to the disc mid-plane and shows increased resolution towards the disc
centre and mid-plane where the density is higher.  A grid with
$f_{\rm{split}}=0.1$, also using $10^7$ particles, is plotted in
Fig.~\ref{fig:density_condition_mesh}. The resolution increases
towards the edge of the disc, where there are large density gradients,
and also increases towards the centre of the disc. 
\begin{figure*}
  \subfigure[AMR grid with a mass per cell condition of
    $5\times10^{26}\,\rm{g}$.]
    {\includegraphics[scale=0.22]{figures/set1_mesh}
  \label{fig:mass_condition_mesh}}
  \subfigure[AMR grid with a density condition of
    $f_{\rm{split}}=0.1$.]
    {\includegraphics[scale=0.22]{figures/set2_mesh}
      \label{fig:density_condition_mesh}}
  \caption{Example AMR grids for a mass per cell splitting condition
    of  $5\times10^{26}\,\rm{g}$ (left)
    and a density contrast splitting condition of $f_{\rm{split}}=0.1$
    (right) constructed
    from $10^7$ SPH particles. The grids are
    slices perpendicular to the disc mid-plane and the axis units are $10^{16}$\,cm.}
  \label{fig:example_meshes}
\end{figure*}

@@

SEDs were calculated using the {\sc torus} code for viewing
angles of 12.5 and 77.5 degrees (where an inclination angle of zero
degrees indicates that the disc is viewed face-on) and the results are
plotted in Fig.~\ref{fig:seds} (dashed line). Also plotted in
Fig.~\ref{fig:seds} are the benchmark results of \cite{pascucci_2004}
(solid line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/sed_013}
  \includegraphics[scale=0.3]{figures/sed_077}
  \caption{SEDs from the benchmark disc represented by $10^5 $ particles and a
    Cartesian radiative transfer grid generated using a mass per cell
    limit of $5\times10^{26}$\,g. SEDs are shown for inclination
    angles of 12.5 degrees (left) and 77.5 degrees (right). The
    benchmark SED is plotted as a solid line and the {\sc torus} SED
    is plotted as a dashed line. The corresponding fractional
    temperature errors are plotted in Fig.\ref{fig:nomod_tem_diff}.}
  \label{fig:seds}
\end{figure*}
Both SEDs show significant departures from the benchmark results;
there is excess emission between 10-100 $\mu\rm{m}$ and at a viewing
angle of 77.5 degrees there is also a significant reduction in flux
below 1\,$\mu\rm{m}$. Both these discrepancies are attributable to a lack of resolution within a few au of the central object. The density distribution in the central region of
this disc is plotted in Fig.~\ref{fig:basic_dens} which shows that the
central 1\,au gap is not represented. There are two effects in
operation; firstly the grid resolution in this region is larger than
1\,au  and secondly the smoothing lengths of the particles are larger
than 1\,au. In order to correctly represent the central gap there needs
to be higher grid resolution around the central source and more SPH
particles are required so that the smoothing lengths are smaller.
\begin{figure}
\includegraphics[scale=0.2]{figures/basic_dens}
\caption{Density in the central region of the benchmark disc
  represented by $10^5$ particles and a Cartesian radiative transfer
  grid generated using a mass per cell limit of
  $5\times10^{26}$\,g. Axis units are $10^{13}$cm. The central 1\,au gap 
   is not represented and the effects on
  the temperature distribution and emergent SEDs are shown in
  Fig.~\ref{fig:nomod_tem_diff} and Fig.~\ref{fig:seds}
  respectively.}
\label{fig:basic_dens}
\end{figure}
